---
title: "group_optimization"
author: "Archana Gadicharla, Nikhila Reddy Vantari"
date: "2024-12-11"
output: html_document
---

## Introduction

This document formulates and solves a mathematical optimization problem to form 4 groups of 3 students each, maximizing the chance of success on a class project based on three defined factors.

## Step 1: Define Factors

We define three factors affecting group success:

1. **GPA**: Higher GPA contributes to project success.
2. **Teamwork Skills**: Rated on a scale from 1 to 5.
3. **Project Experience**: Number of past projects completed.

These factors will be combined into a weighted score to measure each student's potential contribution.


## Step 2: Generate Data

We'll randomly generate data for 12 students for the three factors.

```{r, message=FALSE, warning=FALSE}
set.seed(123)  # For reproducibility

# Number of students
n_students <- 12

# Generate random data for each factor
students <- data.frame(
  ID = 1:n_students,
  GPA = round(runif(n_students, 2.5, 4.0), 2),                # Random GPA between 2.5 and 4.0
  Teamwork_Skills = sample(1:5, n_students, replace = TRUE),  # Random teamwork skills score between 1 and 5
  Project_Experience = sample(0:5, n_students, replace = TRUE) # Random project experience between 0 and 5
)

# Calculate the success score with weights:
# GPA = 0.5, Teamwork Skills = 0.3, Project Experience = 0.2
students$Success_Score <- round(
  students$GPA * 0.5 + students$Teamwork_Skills * 0.3 + students$Project_Experience * 0.2,
  3
)

print(students)
```



## Step 3: Calculate Success Scores

We'll calculate a success score for each student based on weighted factors:

- **GPA Weight**: 0.5  
- **Teamwork Skills Weight**: 0.3  
- **Project Experience Weight**: 0.2  

```{r}
# Define weights
weights <- c(GPA = 0.5, Teamwork_Skills = 0.3, Project_Experience = 0.2)

# Compute success score for each student
students$Success_Score <- with(students, 
  GPA * weights["GPA"] + Teamwork_Skills * weights["Teamwork_Skills"] + Project_Experience * weights["Project_Experience"])

# Display student data with success scores
print(students)
```

## Step 4: Define the Optimization Problem

We want to form 4 groups of 3 students each, maximizing the total group success score.

### Decision Variables

Let \(x_{ij}\) be a binary variable:

- \(x_{ij} = 1\) if student \(i\) is assigned to group \(j\)  
- \(x_{ij} = 0\) otherwise

### Objective Function

Maximize the total success score:

\[
\text{Maximize } \sum_{i=1}^{12} \sum_{j=1}^{4} S_i \cdot x_{ij}
\]

Where \(S_i\) is the success score of student \(i\).

### Constraints

1. Each student is assigned to exactly one group:

   \[
   \sum_{j=1}^{4} x_{ij} = 1 \quad \forall i \in \{1, \ldots, 12\}
   \]

2. Each group has exactly 3 students:

   \[
   \sum_{i=1}^{12} x_{ij} = 3 \quad \forall j \in \{1, \ldots, 4\}
   \]


## Step 5: Solve the Optimization Problem

We'll use the `lpSolve` package to solve the linear programming problem.

```{r, message=FALSE, warning=FALSE}
# Load lpSolve package
library(lpSolve)

# Number of students and groups
n_students <- 12
n_groups <- 4
group_size <- 3

# Calculate success scores with predefined weights
weights <- c(GPA = 0.5, Teamwork_Skills = 0.3, Project_Experience = 0.2)
students$Success_Score <- with(students,
  GPA * weights["GPA"] + Teamwork_Skills * weights["Teamwork_Skills"] + Project_Experience * weights["Project_Experience"]
)

# Display student data
print(students)

# Objective function (flatten the success scores)
success_scores <- as.vector(t(matrix(rep(students$Success_Score, n_groups), nrow = n_groups, byrow = TRUE)))

# Decision variables matrix: x[i, j] where i = student, j = group
# Flattened, this becomes a vector of length n_students * n_groups
n_vars <- n_students * n_groups

# Constraint matrix and RHS
# Each student must be assigned to exactly one group
student_constraints <- matrix(0, nrow = n_students, ncol = n_vars)
for (i in 1:n_students) {
  student_constraints[i, ((i - 1) * n_groups + 1):(i * n_groups)] <- 1
}

# Each group must have exactly 3 students
group_constraints <- matrix(0, nrow = n_groups, ncol = n_vars)
for (j in 1:n_groups) {
  group_constraints[j, seq(j, n_vars, by = n_groups)] <- 1
}

# Combine constraints
constraint_matrix <- rbind(student_constraints, group_constraints)

# Right-hand side constraints
rhs <- c(rep(1, n_students), rep(group_size, n_groups))

# Constraint directions
directions <- c(rep("=", n_students + n_groups))

# Solve the optimization problem
solution <- lp("max", success_scores, constraint_matrix, directions, rhs, all.bin = TRUE)

# Check if the solution is optimal
if (solution$status == 0) {
  cat("Solution found!\n\n")
} else {
  stop("No optimal solution found.")
}

# Display the assignment matrix
assignment <- matrix(solution$solution, nrow = n_students, byrow = TRUE)
colnames(assignment) <- paste("Group", 1:n_groups)
rownames(assignment) <- paste("Student", students$ID)

print(assignment)

# Display group assignments
for (j in 1:n_groups) {
  cat("\n**Group", j, "Members:**\n")
  group_members <- students$ID[which(assignment[, j] == 1)]
  print(students[students$ID %in% group_members, ])
}

```

## Step 6: Group Assignments

Let's display the final group assignments.

```{r}
for (j in 1:n_groups) {
  cat("\n**Group", j, "Members:**\n")
  group_members <- students$ID[which(assignment[, j] == 1)]
  print(students[students$ID %in% group_members, ])
}
```

## Conclusion

We successfully formulated and solved a mathematical optimization problem to form 4 groups of 3 students each, maximizing their potential success based on GPA, teamwork skills, and project experience.